\chapter{Loops}

\section{\texttt{for}}

Iterates over a sequence. The iteration variable is visible inside the
block and \textit{persists} after the loop with the last value it held:

\begin{lstlisting}[caption={for over a range}]
for i in 1..4 {
  display i    // 1, 2, 3  (4 not included)
}
display i    // 3 — variable persists after the loop
\end{lstlisting}

\subsection*{Exclusive and inclusive ranges}

\hayashi{} follows the same convention as Rust:

\begin{lstlisting}
for i in 1..3  { }    // 1, 2      (exclusive: end not included)
for i in 1..=3 { }    // 1, 2, 3   (inclusive: end included)
\end{lstlisting}

Ranges are also expressions that evaluate to lists — see
Section~\ref{sec:ranges}.

\subsection*{Iterating over lists}

\begin{lstlisting}
let fruits = ["apple", "banana", "grape"]

for fruit in fruits {
  display fruit
}
\end{lstlisting}

\subsection*{Discarding the iteration variable}

When the index is not needed:

\begin{lstlisting}
for _ in 1..3 {
  display "repetition"
}
\end{lstlisting}

\section{\texttt{while}}

Executes while the condition is true:

\begin{lstlisting}[caption={Basic while}]
let n = 1

while n <= 5 {
  display n
  n += 1
}
\end{lstlisting}

\begin{aviso}
  \hayashi{} does not detect infinite loops. A \lstinline|while true { }|
  without \lstinline{break} will run indefinitely. Use \texttt{Ctrl-C}
  to interrupt.
\end{aviso}

\section{\texttt{break} and \texttt{continue}}

\lstinline{break} exits the loop immediately. \lstinline{continue} advances
to the next iteration:

\begin{lstlisting}[caption={break and continue}]
for i in 1..10 {
  if i == 3 { continue }    // skip 3
  if i == 7 { break }       // stop at 7
  display i
}
// prints: 1, 2, 4, 5, 6
\end{lstlisting}

\section{Loops with DataFrames}

\lstinline{for} can iterate over lists produced by data functions:

\begin{lstlisting}[caption={Iterating over groups}]
load "sales.csv" as df

for state in ["TX", "CA", "NY"] {
  let sub = filter(df, region == state)
  display f"{state}: {count(sub)} obs"
}
\end{lstlisting}

\section{\texttt{parallel for}}

Concurrent variant of \texttt{for}. Iterations run in parallel across
threads; each iteration's return value (last expression or explicit
\texttt{return}) is collected into a list.

\textbf{As an expression} (preferred --- result goes to a named variable):

\begin{lstlisting}[caption={parallel for as expression}]
let results = parallel for t in tickers {
  let sub = filter(df, ticker == t)
  nrow(sub)
}
// results is a list of nrow values
\end{lstlisting}

\textbf{As a statement} (result stored in the iteration variable):

\begin{lstlisting}[caption={parallel for as statement}]
parallel for t in tickers {
  let sub = filter(df, ticker == t)
  nrow(sub)
}
// t now holds a list of nrow values
\end{lstlisting}

Optional \texttt{threads=N} limits the number of worker threads
(default: all available CPUs):

\begin{lstlisting}[caption={Thread limit}]
let results = parallel for t in tickers, threads=4 {
  load "data.db" as sub, query=f"SELECT * FROM p WHERE t='{t}'"
  nrow(sub)
}
\end{lstlisting}

\subsection*{Isolation model}

Each thread gets its own interpreter with a \textbf{snapshot} of the
outer environment (only send-safe values are captured). This has
important consequences:

\begin{itemize}
  \item \textbf{Reads from outer scope work}: variables defined before
    the \texttt{parallel for} are visible inside the body.
  \item \textbf{Writes to outer scope are silently discarded}:
    reassigning an outer variable inside the body has no effect on the
    outer scope. The mutation lives only in that thread's local
    environment and is lost when the thread exits.
\end{itemize}

\begin{lstlisting}[caption={Mutations are discarded}]
let n = 0
parallel for i in 1..3 {
  n = 10          // no effect on outer n
  i * 2
}
print(n)          // still 0
\end{lstlisting}

\begin{aviso}
  Treat the body as a \textbf{pure function}: read outer variables,
  compute a result, and \texttt{return} it. Do not rely on side effects
  to propagate outward.
\end{aviso}

\subsection*{What can and cannot be captured}

\begin{center}
\begin{tabular}{lll}
\toprule
\textbf{Type} & \textbf{Capturable?} & \textbf{Reason} \\
\midrule
\texttt{Int}, \texttt{Float}, \texttt{Str}, \texttt{Bool} & Yes & Primitive, copied by value \\
\texttt{List}, \texttt{Dict} (send-safe) & Yes & \texttt{Arc}-backed, shared read-only \\
\texttt{DataFrame}, \texttt{Series} & Yes & \texttt{Arc}-backed, operations are functional \\
\texttt{UserFn} & Yes & \texttt{Arc}-backed \\
Model results (\texttt{ols}, \texttt{iv}, \texttt{rolling}, \ldots) & \textbf{No} & Contain \texttt{Rc}, not \texttt{Send} \\
\bottomrule
\end{tabular}
\end{center}

Model results from the outer scope cannot be captured. Compute models
\textbf{inside} the body instead:

\begin{lstlisting}[caption={Models must be computed inside}]
// WRONG: m contains Rc, will be rejected
let m = ols(y ~ x, df)
parallel for i in 1..10 { predict(m, newdata) }

// CORRECT: compute inside the body
parallel for i in 1..10 {
  let sub = filter(df, group == i)
  let m = ols(y ~ x, sub)
  tidy(m)
}
\end{lstlisting}

\subsection*{Skipping iterations and aggregation}

Use \texttt{return nil} to skip an iteration. \texttt{nil} entries are
kept in the result list (in order). Combine with \texttt{rbind()} to
aggregate per-iteration DataFrames --- \texttt{rbind} silently skips
\texttt{nil} entries:

\begin{lstlisting}[caption={parallel for + rbind}]
let results = parallel for i in 0..n, threads=8 {
  let t = tickers[i]
  if t == "SPY" { return nil }
  let df_t = compute_betas(t)
  df_t
}
let all = rbind(results)
\end{lstlisting}
