\chapter{DataFrames}
\label{cap:dataframes}

\section{What is a DataFrame}

A DataFrame is a two-dimensional table with named columns. In \hayashi{},
each column is a 64-bit floating-point vector. Missing values are
represented by \texttt{NaN} (\textit{Not a Number}).

The DataFrame is the central type of the language — virtually every
econometric analysis starts with a DataFrame loaded from some source.

\section{Copy semantics (COW)}

DataFrames use copy-on-write (COW) semantics: assigning a DataFrame to
another variable creates an alias that shares the same internal data. The
actual copy occurs only when one of them is modified.

\begin{lstlisting}[caption={COW: baseline preserved}]
load "panel.csv" as df

let baseline = df      // cheap alias, no copy

let df2 = df |> generate(z = x^2)    // df2 has 'z', df does not
let df3 = df |> filter(year > 2000)  // df3 filtered, df intact

// baseline == original df at any point
\end{lstlisting}

This semantics allows creating analysis variants with no memory cost until
an actual modification is needed.

\section{Creating DataFrames}

\subsection*{\texttt{input} block}

For small inline datasets — ideal for examples and tests:

\begin{lstlisting}[caption={input block}]
input x y z
  1   2  10
  3   4  20
  5   6  30
end
\end{lstlisting}

The \lstinline{input} block creates a DataFrame named \lstinline{df} by
default. The number of columns is defined by the header; each data row
must have the same number of values.

\begin{nota}
  \lstinline{input} accepts only numeric values. Use \lstinline{.} to
  represent missing values. For data with text columns, use
  \lstinline{load}.
\end{nota}

\subsection*{\texttt{dataframe(n)} — pre-dimensioned empty DataFrame}

Create an empty DataFrame with a fixed number of rows. This is useful
before populating it with \lstinline{generate}:

\begin{lstlisting}[caption={Pre-dimensioned DataFrame}]
let df = dataframe(1000)

generate df u = rnormal(0, 1)
generate df treat = rbernoulli(0.5)
\end{lstlisting}

\subsection*{\texttt{load} — external files}

\begin{lstlisting}[caption={Loading files}]
load "data.csv" as sales
load "panel.dta" as panel        // Stata format
load "results.parquet" as df
load "sheet.xlsx" as df          // first sheet
load "sheet.xlsx" sheet="Data" as df
load "base.db" as df, table=prices
load "base.db" as df, query="SELECT * FROM prices WHERE year > 2020"
load "data.csv" as df, sep=";"
\end{lstlisting}

Supported formats: CSV, TSV, JSON, Parquet, Stata (\texttt{.dta}),
Excel (\texttt{.xlsx}/\texttt{.xls}/\texttt{.ods}), SQLite, ODBC. HTTP(S) URLs
are downloaded automatically.

\subsection*{Column projection and row filtering (\texttt{columns=}, \texttt{where=})}

For large files, loading every column and row into RAM can be wasteful or
impossible. The \texttt{columns=} and \texttt{where=} options push projection
and filtering down to the data source: only the requested columns and the
rows matching the predicate are materialized in memory.

\begin{lstlisting}[caption={Column and filter pushdown}]
load "quotes.parquet" as aapl, columns=[ticker, date, adj_close], where="ticker == \"AAPL\""
load "base.db" as df, table=prices, columns=[date, close], where="close > 100"
load "survey.csv" as df, where="age >= 18 && region == \"South\""
load "panel.dta" as df, columns=[id, year, y], where="year >= 2000"
load "sheet.xlsx" as df, columns=[name, score], where="score > 75"
\end{lstlisting}

\begin{center}
\begin{tabular}{llll}
\toprule
\textbf{Source} & \textbf{\texttt{columns=}} & \textbf{\texttt{where=}} & \textbf{Pushdown mechanism} \\
\midrule
Parquet   & yes & yes & Arrow \texttt{ProjectionMask} + \texttt{RowFilter} + \textit{row group pruning} by min/max \\
SQLite    & yes & yes & \texttt{SELECT cols FROM t WHERE pred} \\
ODBC      & yes & yes & same as SQLite \\
CSV / TSV & yes & yes & projection in read loop, row-by-row predicate \\
DTA       & yes & yes & projection in \texttt{read\_record}, row-by-row predicate \\
Excel     & yes & yes & projection after \texttt{worksheet\_range}, cell filter \\
JSON      & not yet & not yet & --- \\
\bottomrule
\end{tabular}
\end{center}

Passing \texttt{columns=} or \texttt{where=} to a JSON \texttt{load} returns
an error. \texttt{query=} cannot be combined with \texttt{columns=} or
\texttt{where=} --- if you need custom SQL, write the projection and filter
inside the \texttt{query=} string itself.

\subsubsection*{\texttt{columns=} syntax}

Accepts a list of column names (identifiers or string literals):

\begin{lstlisting}
columns=[ticker, date, adj_close]
columns=["ticker", "date", "adj_close"]
columns=ticker              // single column also accepted
\end{lstlisting}

Names are validated against the source schema before any data is read;
unknown columns produce an error listing the available ones.

\subsubsection*{\texttt{where=} syntax}

\texttt{where=} accepts a Hayashi expression of the form \texttt{column OP literal},
combined with \texttt{\&\&}, \texttt{||}, \texttt{!} and \texttt{in [...]}.
The expression is parsed by the Hayashi parser and normalized into a
\texttt{RowPredicate} that each loader evaluates (or pushes down to the source).

\begin{lstlisting}
where="ticker == \"AAPL\""                     // string equality
where="price > 100"                            // numeric comparison
where="price > 100 && volume > 1e6"            // AND
where="ticker in [\"AAPL\", \"MSFT\"]"         // membership
where="!(sector == \"Finance\")"               // negation
where="year >= 2022 && product == \"Soy\""     // combined
\end{lstlisting}

Supported operators: \texttt{==}, \texttt{!=}, \texttt{>}, \texttt{>=},
\texttt{<}, \texttt{<=}, \texttt{in}, \texttt{\&\&} (and), \texttt{||} (or),
\texttt{!} (not). Comparisons must be \texttt{column OP literal} --- comparing
two columns is not supported (use \texttt{generate} + \texttt{filter} for that).

\begin{nota}
  On large Parquet files, filter pushdown is especially effective: the
  predicate is evaluated during the row-group scan, before any data is decoded
  into memory. On an 837\,MB file with 30 million rows and 8{,}292 tickers,
  loading with
  \texttt{columns=[ticker, date, adj\_close], where="ticker == \textbackslash"AAPL\textbackslash""}
  reduced peak RAM from $\sim$7.5\,GB (full eager load) to $\sim$4\,MB.
\end{nota}

\subsubsection*{Row group pruning by statistics (Parquet)}

In addition to \texttt{RowFilter} (which evaluates the predicate on each row
during scan), the Parquet loader reads per-row-group min/max statistics from
the file metadata and skips entire row groups where the predicate cannot
possibly match. This is especially effective when the data is sorted or
clustered by the filtered column.

On the same 799\,MB file with 30 million rows and 8{,}292 row groups sorted by
ticker, \texttt{where="ticker == \textbackslash"AAPL\textbackslash""} pruned
8{,}291 of 8{,}292 row groups, reducing load time from $\sim$62\,s (full scan)
to $\sim$312\,ms (212$\times$ faster) with $\sim$60\,MB peak RAM.

\begin{nota}
  For point lookups (fetching a single value, such as one specific ticker),
  SQLite with a B-tree index on \texttt{(ticker, date)} is still faster
  ($\sim$42\,ms, $\sim$26\,MB RAM) because it seeks directly without reading
  all row-group metadata. For full-column analytics (all rows, few columns),
  Parquet with pruning is superior due to columnar compression.
\end{nota}

\subsubsection*{Temporal types in Parquet}

Columns of type \texttt{Timestamp(s|ms|$\mu$s|ns)}, \texttt{Date32}, and
\texttt{Date64} in Parquet are converted to \textbf{ISO 8601 strings}:
\texttt{YYYY-MM-DD} when the time component is midnight, or
\texttt{YYYY-MM-DDTHH:MM:SS} otherwise. Hayashi does not have a native
\textit{Date} type, so the textual representation is the most practical for
econometrics --- it allows component extraction via \texttt{substr} and
conversion to a Unix timestamp when needed.

\begin{lstlisting}[caption={Extracting date components}]
load "cotacoes.parquet" as df, columns=[ticker, date, close]

generate df year = substr(date, 0, 4)       // "2024"  (vectorial)
generate df month = substr(date, 5, 2)      // "03"    (vectorial)
\end{lstlisting}

\begin{nota}
  \texttt{substr()} is vectorial and works inside \texttt{generate}, producing
  string columns. In contrast, \texttt{date()} and \texttt{datetime()} are
  scalar functions --- they accept a single ISO string and return a \texttt{Float}
  (seconds since 1970-01-01). Example: \texttt{let ts = date("2024-03-15")}.
\end{nota}

\section{Saving DataFrames}

\begin{lstlisting}
save df "output.csv"
save df "output.parquet"
save df "output.dta"
save df "output.xlsx"
\end{lstlisting}

\section{Inspecting DataFrames}

\subsection*{\texttt{display}}

Displays the first rows of the DataFrame:

\begin{lstlisting}
display df
display df, 20    // display up to 20 rows
\end{lstlisting}

\subsection*{\texttt{describe}}

Displays the structure: number of observations, variables, and types:

\begin{lstlisting}
describe df
\end{lstlisting}

Typical output:

\begin{lstlisting}[style=inline]
DataFrame: 1250 obs, 8 vars
  year    Float  (0 missing)
  state   Float  (0 missing)
  income  Float  (12 missing)
  gini    Float  (3 missing)
\end{lstlisting}

\subsection*{\texttt{count} and \texttt{nrow}}

Returns the number of rows:

\begin{lstlisting}
let n  = count(df)               // total observations
let n2 = nrow(df)                // alias for count(df)
let n3 = count(df, income > 1000) // conditional count
\end{lstlisting}

\section{Special variables}

Inside row-by-row commands (\lstinline{generate}, \lstinline{mutate},
\lstinline{filter}), two implicit variables are available:

\begin{description}
  \item[\lstinline{_n}] Current row number (starts at 1).
  \item[\lstinline{_N}] Total rows in the DataFrame.
\end{description}

\begin{lstlisting}
generate df id     = _n       // 1, 2, 3, ..., N
generate df weight = 1.0/_N   // uniform weight
\end{lstlisting}

\section{Time series operators}

Inside \lstinline{generate} expressions (imperative form), special
time-series operators are available:

\begin{center}
\begin{tabular}{cll}
\toprule
\textbf{Operator} & \textbf{Meaning} & \textbf{Example} \\
\midrule
\texttt{L.x}   & Lag (previous value)   & \texttt{L.gdp}  → gdp$_{t-1}$ \\
\texttt{F.x}   & Lead (next value)      & \texttt{F.rate} → rate$_{t+1}$ \\
\texttt{D.x}   & First difference       & \texttt{D.gdp}  → gdp$_t$ - gdp$_{t-1}$ \\
\texttt{L2.x}  & Lag of order 2         & \texttt{L2.gdp} → gdp$_{t-2}$ \\
\bottomrule
\end{tabular}
\end{center}

\begin{lstlisting}[caption={Time series operators}]
generate df dgdp    = D.gdp     // GDP growth
generate df gdp_lag = L.gdp     // lagged GDP
\end{lstlisting}
