\chapter{Modules, Packages, and Plugins}

\section{The Hayashi Module System}

In \hayashi{}, code sharing and organization are handled via two main commands: \lstinline{source} and \lstinline{import}.

\subsection{The \texttt{source} command}
The \lstinline{source} command executes a \texttt{.hay} file directly within the current context. All definitions (functions and variables) inside the file become visible in the global scope of the caller, without namespaces:

\begin{lstlisting}[caption={Basic usage of source}]
source "lib/utils.hay"
let result = my_utility_function(10)
\end{lstlisting}

\subsection{The \texttt{import} command}
The \lstinline{import} command loads a module and isolates its functions under a **namespace** using the \texttt{::} delimiter. This prevents naming conflicts and keeps your code organized:

\begin{lstlisting}[caption={Importing with namespaces}]
// Import and use the filename as the default namespace
import("lib/math.hay")
let r1 = math::compute_mean([10, 20, 30])

// Import using a custom alias
import("lib/math.hay", as=m)
let r2 = m::compute_mean([10, 20, 30])

// Import specific functions directly into the local scope
import("lib/math.hay", only=["compute_mean"])
let r3 = compute_mean([10, 20, 30])
\end{lstlisting}

---

\section{Package Manager and GitHub Publishing}

\hayashi{} uses **GitHub directly as its package registry**. This eliminates the need for a proprietary central registry and promotes decentralized open-source package distribution.

\subsection{CLI Commands}
To manage third-party packages from your operating system terminal, use:

\begin{lstlisting}[style=inline, numbers=none, frame=none]
hay install usuario/repositorio   -- Install a package from GitHub (-y to skip overwrite prompt)
hay remove  usuario/repositorio   -- Uninstall a package (deletes script dirs, native files, and metadata)
hay list                         -- List installed packages and plugins
hay check-plugin [user/repo]     -- Check integrity/version with remote repository
hay update [user/repo] [-y]      -- Update package(s) to latest version (-y to skip prompt)
\end{lstlisting}

\subsection{Standard Package Structure}

Hayashi packages are convention-based and require zero configuration:
\begin{itemize}
  \item \textbf{Script Packages}: Simply push your \texttt{.hay} script(s) to the root of a GitHub repository (e.g. \texttt{github.com/john/finance/interest.hay}).
  \item \textbf{Native Plugins}: Set up a GitHub Actions workflow to build and upload compiled binaries (\texttt{.so}, \texttt{.dll}, \texttt{.dylib}, or \texttt{.wasm}) as assets to a GitHub Release.
\end{itemize}

Once your package is pushed to GitHub, any user in the world can install and use it in Hayashi, either from the CLI or directly inside a script:
\begin{lstlisting}[caption={Installing and importing a package}]
// In the OS terminal:
// hay install john/finance

// Inside the Hayashi script itself:
install("john/finance")
import("john/finance/interest")
let rate = finance::compute_rate(1000, 12)
\end{lstlisting}

The \lstinline{install("user/repo")} function also accepts the optional arguments \lstinline{version} (for a specific version, e.g. \lstinline{version="v1.2.3"}) and \lstinline{force=true} (to overwrite an existing installation without prompting).

---

\section{Developing Advanced Plugins}

For needs demanding high performance or integration with other languages, Hayashi supports three plugin types under the exact same \lstinline{import} syntax:

\subsection{1. Script Plugins (.hay)}
Written in pure Hayashi. Ideal for macro-like econometric routines and fast utilities.

\subsection{2. Native Plugins (Exclusive to Rust)}
Native plugins are developed in Rust to guarantee maximum performance and memory safety. They communicate with Hayashi through a stable, decoupled C-string JSON ABI.

To make development easier, you should use the official \texttt{hayashi-plugin-sdk}. The SDK provides proc-macros to automatically generate all unsafe FFI boilerplate and handle JSON serialization/deserialization:

\begin{lstlisting}[caption={Developing a native plugin using the SDK}]
use hayashi_plugin_sdk::{hayashi_fn, hayashi_plugin};

// The #[hayashi_fn] macro automatically handles JSON (de)serialization.
// You can use standard Rust types (Vec, HashMap, String, primitives, etc.).
#[hayashi_fn]
pub fn distance_to_point(lats: Vec<f64>, lons: Vec<f64>, ref_lat: f64, ref_lon: f64) -> Vec<f64> {
    lats.iter()
        .zip(lons.iter())
        .map(|(&lat, &lon)| {
            // haversine distance calculation
            let r = 6371.0;
            let d_lat = (ref_lat - lat).to_radians();
            let d_lon = (ref_lon - lon).to_radians();
            let a = (d_lat / 2.0).sin().powi(2)
                + lat.to_radians().cos() * ref_lat.to_radians().cos() * (d_lon / 2.0).sin().powi(2);
            let c = 2.0 * a.sqrt().atan2((1.0 - a).sqrt());
            r * c
        })
        .collect()
}

// Generates the required FFI memory cleanup functions
hayashi_plugin!();
\end{lstlisting}

Compiled as a \texttt{cdylib}, you can publish your plugin by creating a tag (e.g. \texttt{v0.1.0}) on GitHub. The GitHub Actions release workflow compiles the plugin for Linux (\texttt{.so}), macOS (\texttt{.dylib}), and Windows (\texttt{.dll}), uploading them as release assets.

Any user can install it from the CLI or inside a script:
\begin{lstlisting}[style=inline, numbers=none, frame=none]
hay install john/spatial     // from the OS terminal
install("john/spatial")      // inside a Hayashi script
\end{lstlisting}
The package manager automatically downloads the binary that matches the current operating system and architecture.

You can then load it and name its namespace in Hayashi:
\begin{lstlisting}
import("john/spatial", as=geo)
let dist = geo::distance_to_point(df["lat"], df["lon"], -30.03, -51.22)
generate df dist_poa = dist
\end{lstlisting}

\subsubsection{Security and Memory Considerations: JSON ABI vs. Zero-Copy FFI}
While the default C-string JSON ABI offers extreme safety and decoupling (as any schema evolution in JSON doesn't break memory layouts), it can become a bottleneck when sharing large volumes of data (Big Data) due to serialization overhead. 

To bridge this gap, Hayashi supports transparent zero-copy memory sharing using the Apache Arrow FFI format for complex array/dataframe exchanges. This is implemented automatically by passing pointers to FFI structs (\texttt{FFI\_ArrowArray} and \texttt{FFI\_ArrowSchema}) in a zero-copy manner. However, developers must find the right balance:
\begin{itemize}
    \item \textbf{JSON C String}: Offers maximum decoupling, safety, and binary version compatibility across different compiler versions. Best for control routines, small datasets, and general extensibility.
    \item \textbf{Apache Arrow Zero-Copy}: Offers maximum speed and zero memory allocation overhead. Best for data-intensive calculations, deep learning inference, or large econometric estimations. It requires strict memory lifecycle coordination (manual prevention of double-free bugs on the FFI boundary).
\end{itemize}

This duality represents an intentional design decision of the Hayashi language. Restricting zero-copy support strictly to DataFrames and homogeneous columns protects the ecosystem from binary compatibility (ABI) breakage in complex control structures and metadata, which are cheap to serialize via JSON C String. This guarantees long-term stability and ease of maintenance of the language without sacrificing performance on large data volume exchanges.


\subsection{3. WebAssembly Plugins (WASM - For other languages)}
For developers using other languages (such as C, Go, Zig, etc.), plugins must be compiled to WebAssembly (\texttt{.wasm}). They run inside a fully sandboxed environment via Hayashi's built-in WASM engine, ensuring absolute safety (any crash inside the plugin will not crash Hayashi itself) and instant cross-platform portability.

The WASM plugin must export \texttt{alloc(size: i32) -> i32} and \texttt{dealloc(ptr: i32, size: i32)} functions to manage passing JSON strings into WASM linear memory, and the plugin function should return an \texttt{i64} representing the pointer (in the upper 32 bits) and length (in the lower 32 bits) of the returned JSON string.
