\chapter{GARCH and Conditional Volatility}
\label{cap:garch}
% ── index ──────────────────────────────────────────────────────
\index{GARCH|textbf}
\index{conditional volatility|textbf}
\index{conditional heteroskedasticity|textbf}
\index{ARCH|see{GARCH}}
\index{garch@\texttt{garch()}|textbf}
\index{volatility clustering}
\index{EGARCH}
\index{GJR-GARCH}

% ─────────────────────────────────────────────────────────────────────────────
\section{Conditional Heteroskedasticity}

Financial series exhibit a recurring phenomenon: periods of high
volatility tend to be followed by more volatility, and calm periods
persist. This \emph{volatility clustering} violates the assumption of
constant conditional variance in ARMA models — the variance of $\varepsilon_t$
given the past varies over time.

The GARCH(\textit{p,q}) model (Engle 1982; Bollerslev 1986) captures this
dynamic explicitly. The GARCH(1,1) specification is:

\[
  \varepsilon_t \mid \mathcal{F}_{t-1} \sim (0,\, h_t)
  \qquad
  h_t = \omega + \alpha_1\,\varepsilon_{t-1}^2 + \beta_1\,h_{t-1}
\]

\noindent where $h_t$ is the \emph{conditional variance}, $\alpha_1$ captures the
ARCH effect (reaction to past shock) and $\beta_1$ captures persistence
(volatility memory).

\begin{center}
\begin{tabular}{lll}
\toprule
\textbf{Restriction} & \textbf{Condition} & \textbf{Meaning} \\
\midrule
Positive variance   & $\omega > 0,\; \alpha_1 \geq 0,\; \beta_1 \geq 0$ & $h_t > 0$ guaranteed \\
Stationarity        & $\alpha_1 + \beta_1 < 1$                           & variance does not explode \\
Unconditional variance & $\sigma^2 = \omega/(1-\alpha_1-\beta_1)$       & long-run level \\
\bottomrule
\end{tabular}
\end{center}

The case $\alpha_1 + \beta_1 = 1$ is called IGARCH (\emph{integrated}): the
volatility has a unit root and shocks have a permanent effect.

% ─────────────────────────────────────────────────────────────────────────────
\section{ARCH-LM Test}

Before estimating a volatility model, one tests whether ARCH effects are
present. Engle's (1982) test regresses $\varepsilon_t^2$ on its own lags and
evaluates joint significance:

\[
  \varepsilon_t^2 = a_0 + a_1\,\varepsilon_{t-1}^2 + \cdots + a_m\,\varepsilon_{t-m}^2 + u_t
\]

\[
  H_0{:}\; a_1 = \cdots = a_m = 0
  \quad \text{(no ARCH effects)}
\]

The statistic is $LM = N \cdot R^2_{\text{aux}} \xrightarrow{d} \chi^2(m)$
under $H_0$.

% ─────────────────────────────────────────────────────────────────────────────
\section{DGP Verification}

The DGP generates a true ARCH(1) process ($\beta=0$):

\[
  h_t = 0{,}3 + 0{,}5\,\varepsilon_{t-1}^2
  \qquad
  \varepsilon_t = \sqrt{h_t}\; z_t
  \qquad
  z_t \sim (0,1)
\]

The innovation is generated directly as a standard normal draw:
$z = \text{rnormal}()$, so $E[z]=0$ and $E[z^2]=1$.

\begin{lstlisting}[caption={ARCH(1) DGP and GARCH estimation}]
seed(42)
input df
id
1
end
for i in 1..=9 { df = append(df, df) }
df |> filter(_n <= 500)
generate df z  = rnormal()               // z: E[z]=0, E[z²]=1
generate df e  = z
generate df t  = _n
let n = nrow(df)
let i = 2
while i <= n {
    let le = mean(df, e, if = t == i - 1)
    let lz = mean(df, z, if = t == i)
    replace df e = sqrt(0.3 + 0.5 * le * le) * lz if t == i
    i = i + 1
}

archtest(df, e, lags=5)      // should reject H₀
let m = garch(df, e, p=1, q=1)
display m
archtest(m, lags=5)          // should not reject (GARCH captured the ARCH)
\end{lstlisting}

\subsection*{ARCH-LM Test — raw series}

The raw series should reject the no-ARCH null, confirming time-varying
conditional variance in the simulated process.

\subsection*{GARCH(1,1) Estimation}

The MLE estimator recovers the true DGP ($\omega=0{,}3$, $\alpha=0{,}5$,
$\beta=0$) up to finite-sample variation. A fitted $\beta$ near zero identifies
that the process is pure ARCH — with no GARCH persistence component.\footnote{%
  With $N=500$, finite-sample estimates need not match the DGP parameters
  exactly; with larger $N$, convergence improves.}

\subsection*{ARCH-LM on standardized residuals}

After fitting the GARCH, the standardized residuals $\hat{z}_t = \varepsilon_t / \sqrt{\hat{h}_t}$
should be white noise in $z_t^2$:

After fitting the model, the ARCH-LM test on standardised residuals should no
longer reject if the volatility specification has captured the conditional
heteroskedasticity.

% ─────────────────────────────────────────────────────────────────────────────
\section{Variants: EGARCH and GJR-GARCH}

The symmetric GARCH treats positive and negative shocks equally. In
financial assets, negative shocks tend to increase volatility more
than positive shocks of the same magnitude (\emph{leverage effect}).

\textbf{EGARCH} (Nelson 1991) models $\ln h_t$:
\[
  \ln h_t = \omega + \beta_1 \ln h_{t-1}
             + \alpha_1 \left|\frac{\varepsilon_{t-1}}{\sqrt{h_{t-1}}}\right|
             + \gamma_1 \frac{\varepsilon_{t-1}}{\sqrt{h_{t-1}}}
\]
The parameter $\gamma_1 < 0$ captures asymmetry (leverage effect).

\textbf{GJR-GARCH} (Glosten-Jagannathan-Runkle 1993) adds an indicator term:
\[
  h_t = \omega + \alpha_1 \varepsilon_{t-1}^2
        + \gamma_1 \varepsilon_{t-1}^2 \mathbf{1}[\varepsilon_{t-1} < 0]
        + \beta_1 h_{t-1}
\]
The coefficient $\gamma_1 > 0$ implies that negative shocks raise volatility more.

% ─────────────────────────────────────────────────────────────────────────────
\section{Syntax}

\begin{lstlisting}
garch(df, y, p=1, q=1)
garch(df, y, p=1, q=1, dist=t)    // Student-t errors
egarch(df, y, p=1, q=1)
gjrgarch(df, y, p=1, q=1)
archtest(df, y, lags=5)            // ARCH-LM on raw series
archtest(modelo, lags=5)           // ARCH-LM on standardized residuals
\end{lstlisting}

\begin{center}
\begin{tabular}{lll}
\toprule
\textbf{Parameter} & \textbf{Default} & \textbf{Description} \\
\midrule
\texttt{p}    & 1                & ARCH order ($\alpha$) \\
\texttt{q}    & 1                & GARCH order ($\beta$) \\
\texttt{dist} & \texttt{normal}  & distribution: \texttt{normal} or \texttt{t} \\
\texttt{lags} & 5                & lags for ARCH-LM \\
\bottomrule
\end{tabular}
\end{center}

\begin{lstlisting}[caption={Typical workflow — asset returns}]
load "retornos.csv" as df

// 1. Check for volatility clustering
archtest(df, ret, lags=10)

// 2. Estimate volatility model
let m_garch  = garch(df, ret, p=1, q=1)
let m_gjr    = gjrgarch(df, ret, p=1, q=1)

// 3. Check diagnostics
archtest(m_garch, lags=10)

// 4. Estimated conditional variance
display m_garch
\end{lstlisting}

\begin{nota}
  \hayashi{} estimates GARCH by MLE assuming Gaussian innovations
  (\lstinline{dist=normal}) or Student-t (\lstinline{dist=t}). For series
  with very heavy tails (cryptocurrencies, exchange rates in crises), the
  $t$ distribution generally improves the fit. Compare AIC/BIC across specifications.
\end{nota}

\section{Empirical validation}

The ARCH(1) DGP in this chapter corresponds to the \texttt{garch\_book}
validation case in \texttt{validation/cases/}. The case compares \hayashi{}'s
MLE coefficients with the Python \texttt{arch} reference. Run
\lstinline{hay validate} to see the current status.
