\chapter{Survival Analysis}
\label{cap:sobrevivencia}
% ── index ──────────────────────────────────────────────────────
\index{survival analysis|textbf}
\index{Kaplan-Meier, estimator|textbf}
\index{km@\texttt{km()}|textbf}
\index{Cox, model|textbf}
\index{cox@\texttt{cox()}|textbf}
\index{hazard function|textbf}
\index{hazard ratio|textbf}
\index{censoring}
\index{proportional hazards, assumption of}

% ─────────────────────────────────────────────────────────────────────────────
\section{Failure Times and Censoring}

Survival analysis models the \emph{time to an event} $T$ —
firm bankruptcy, death, school dropout. The central challenge is
\textbf{right censoring}: when the observation ends before the event,
we only know $T > c$ (observation time $c$), not the exact time.

Two fundamental concepts:

\[
  S(t) = P(T > t)
  \quad\text{(survival function)}
  \qquad
  h(t) = \lim_{\delta\to 0} \frac{P(t \leq T < t+\delta \mid T \geq t)}{\delta}
  \quad\text{(hazard function)}
\]

$S(t)$ decreases from 1 to 0; $h(t)$ measures the instantaneous failure rate
given that the individual survived to $t$.

\subsection*{Kaplan-Meier}

The Kaplan-Meier (1958) non-parametric estimator estimates $S(t)$ without
distributional assumptions:

\[
  \hat{S}(t) = \prod_{t_j \leq t} \left(1 - \frac{d_j}{n_j}\right)
\]

where $d_j$ is the number of events at time $t_j$ and $n_j$ the number at risk
immediately before.

\subsection*{Cox — Proportional Hazard}

The Cox (1972) model specifies:

\[
  h(t \mid x_i) = h_0(t)\,\exp(\beta'\,x_i)
\]

The baseline hazard function $h_0(t)$ is left non-parametric. The
coefficients $\beta$ are estimated by partial likelihood, without
specifying $h_0$. The exponential $e^{\beta_j}$ is the \emph{hazard ratio}:
a unit increase in $x_j$ multiplies the hazard by $e^{\beta_j}$.

% ─────────────────────────────────────────────────────────────────────────────
\section{DGP Verification}

The DGP uses Weibull failure times with shape parameter 1{,}5 and
proportional hazard in $x \in \{0,1\}$. The true log-hazard ratio is $\beta_x = 0{,}8$.

\begin{lstlisting}[caption={Survival DGP — Weibull with proportional hazard}]
seed(42)
generate df x      = rbernoulli(0.5)
generate df u      = runiform()
generate df t_true = exp(-ln(u) / exp(0.8 * x))   // Weibull, beta=0.8
generate df c      = 3.0
generate df time   = t_true * (t_true <= c) + c * (t_true > c)
generate df event  = (t_true <= c)

let m_km  = km(time, event, df)
display m_km

let m_cox = cox(time ~ x, df, event=event)
display m_cox
\end{lstlisting}

\subsection*{Kaplan-Meier}

\begin{lstlisting}[language={}, basicstyle=\ttfamily\small,
  frame=none, numbers=none, backgroundcolor=\color{codbg},
  xleftmargin=0pt, xrightmargin=0pt, breaklines=false]
============== Kaplan-Meier Survival Estimate ==============
Observations:               300
Events:                     246
Median survival:         1.6319
\end{lstlisting}

The median survival is 1{,}63: half of the individuals experience the
event before this time.

\subsection*{Cox}

\begin{lstlisting}[language={}, basicstyle=\ttfamily\small,
  frame=none, numbers=none, backgroundcolor=\color{codbg},
  xleftmargin=0pt, xrightmargin=0pt, breaklines=false]
======================= Cox Proportional Hazards Model =======================
Observations: 300   Events: 246   Log-Likelihood: -1235.9213
Concordance:  0.6725
------------------------------------------------------------------------------
Variable  |    coef |  std err |       z |  P>|z| | exp(coef)
x         |  0.7018 |   0.1306 |   5.372 |  0.000 |    2.0174
\end{lstlisting}

$\hat\beta_x = 0{,}702$ (true: 0{,}8) — within the confidence interval.
The estimated hazard ratio $e^{0{,}702} = 2{,}02$: individuals with $x=1$ have
an instantaneous hazard twice as large as $x=0$. The concordance index
$C = 0{,}67$ indicates moderate discrimination (0{,}5 = random, 1{,}0 = perfect).

% ─────────────────────────────────────────────────────────────────────────────
\section{Syntax}

\begin{lstlisting}
km(time, event, df)                          // Kaplan-Meier
cox(time ~ x1 + x2, df, event=event_col)    // Cox PH
\end{lstlisting}

The \lstinline{event} column must be $1$ for event and $0$ for censoring.

\begin{nota}
  The Cox model assumes \emph{proportional hazards}: $h(t \mid x=1)/h(t \mid x=0)$
  is constant over time. Violations can be detected via Schoenfeld residuals
  or interactions with $\ln(t)$.
\end{nota}
