\chapter{Generalized Linear Models}
\label{cap:glm}
% ── index ───────────────────────────────────────────────────────
\index{generalized linear models|textbf}
\index{GLM|see{generalized linear models}}
\index{glm@\texttt{glm()}|textbf}
\index{PPML|textbf}
\index{link function|textbf}
\index{gamma, distribution}
\index{generalized estimating equations}

% ─────────────────────────────────────────────────────────────────────────────
\section{The exponential family}

Generalized Linear Models (Nelder and Wedderburn 1972) unify OLS,
Logit, Poisson and many other estimators under a common framework. Three
components define a GLM:

\begin{enumerate}
  \item \textbf{Random component}: $Y_i \sim$ exponential family with
  mean $\mu_i$.
  \item \textbf{Linear predictor}: $\eta_i = x_i'\beta$.
  \item \textbf{Link function}: $g(\mu_i) = \eta_i$, connecting the mean
  to the linear scale.
\end{enumerate}

\begin{center}
\begin{tabular}{llll}
\toprule
\textbf{Family} & \textbf{Default link} & \textbf{Variance} & \textbf{Application} \\
\midrule
Gaussian    & identity   & $\sigma^2$   & OLS \\
Binomial    & logit      & $\mu(1-\mu)$ & binary data \\
Poisson     & log        & $\mu$        & count data \\
Gamma       & log        & $\mu^2/\phi$ & positive data with heavy tails \\
\bottomrule
\end{tabular}
\end{center}

Estimation by IRLS (Iteratively Reweighted Least Squares) is efficient and
converges quickly.

\subsection*{GLM Gamma — positive data with heavy tails}

Expenditures, loan sizes and durations follow right-skewed distributions
with variance proportional to the square of the mean. The Gamma GLM with
log link models $E[Y \mid x] = \exp(x'\beta)$ with $\mathrm{Var}[Y\mid x] \propto \mu^2$.

\subsection*{PPML — international trade}

The PPML (Poisson Pseudo MLE) estimator by Santos Silva and Tenreyro (2006) uses
the Poisson estimating equation even when $Y$ is not an integer (e.g., trade values).
It is robust to zeros and to multiplicative heteroscedasticity —
common problems in gravity equations of trade.

% ─────────────────────────────────────────────────────────────────────────────
\section{DGP Verification}

\begin{lstlisting}[caption={GLM Gamma and PPML}]
seed(42)
generate df x  = rnormal()
generate df mu = exp(0.5 + 1.0 * x)
// Gamma: mean=mu, variance ~ mu² (simulated via lognormal)
generate df y_gamma = mu * exp(rnormal() * 0.6)

let m_gamma = glm(y_gamma ~ x, df, family=gamma, link=log)
display m_gamma

// PPML with "trade" data (mix of zeros and positives)
generate df x2    = rnormal()
generate df trade = round(exp(0.3 + 1.2*x + 0.8*x2) * (runiform() > 0.2))
let m_ppml = glm(trade ~ x + x2, df, family=poisson, link=log)
display m_ppml
\end{lstlisting}

\subsection*{GLM Gamma}

The Gamma model should recover positive intercept and slope estimates close to
the log-mean DGP, with dispersion summarising the remaining relative variance.

\subsection*{PPML}

PPML should recover positive effects for both covariates while handling zeros
in the \texttt{trade} variable.

% ─────────────────────────────────────────────────────────────────────────────
\section{Syntax}

\begin{lstlisting}
glm(y ~ x, df, family=gaussian, link=identity)  // OLS via GLM
glm(y ~ x, df, family=binomial, link=logit)     // Logit via GLM
glm(y ~ x, df, family=gamma,    link=log)       // Gamma with log link
glm(y ~ x, df, family=poisson,  link=log)       // Poisson / PPML
\end{lstlisting}

\begin{center}
\begin{tabular}{ll}
\toprule
\textbf{Family} & \textbf{Available links} \\
\midrule
\texttt{gaussian}  & \texttt{identity}, \texttt{log}, \texttt{inverse} \\
\texttt{binomial}  & \texttt{logit}, \texttt{probit}, \texttt{cloglog} \\
\texttt{poisson}   & \texttt{log}, \texttt{identity}, \texttt{sqrt} \\
\texttt{gamma}     & \texttt{log}, \texttt{identity}, \texttt{inverse} \\
\bottomrule
\end{tabular}
\end{center}
