\chapter{Time Series Filters and Decomposition}
\label{cap:filtros}
% ── index ───────────────────────────────────────────────────────
\index{decomposition filters|textbf}
\index{Hodrick-Prescott, filter of|textbf}
\index{hprescott@\texttt{hprescott()}|textbf}
\index{Baxter-King, filter of|textbf}
\index{baxter_king@\texttt{baxter\_king()}|textbf}
\index{Holt-Winters|textbf}
\index{holtwinters@\texttt{holtwinters()}|textbf}
\index{ETS (Error-Trend-Seasonal)|textbf}
\index{trend}
\index{cycle}
\index{seasonality}

% ─────────────────────────────────────────────────────────────────────────────
\section{Separation of components}

A time series can be decomposed into:
\[
  y_t = \tau_t + c_t + s_t + \varepsilon_t
\]
where $\tau_t$ is the trend, $c_t$ the cyclical component, $s_t$ the
seasonality and $\varepsilon_t$ the idiosyncratic noise. Linear filters
extract one or more of these components from the observed series.

\subsection*{Hodrick-Prescott Filter (HP)}

The HP filter (Hodrick and Prescott 1997) separates trend from cycle by minimizing:

\[
  \min_{\{\tau_t\}} \left\{ \sum_{t=1}^T (y_t - \tau_t)^2
  + \lambda \sum_{t=2}^{T-1} [(\tau_{t+1} - \tau_t) - (\tau_t - \tau_{t-1})]^2 \right\}
\]

The parameter $\lambda$ penalizes changes in the second difference of the trend.
Canonical values: $\lambda = 1600$ (quarterly), $100$ (annual), $14400$ (monthly).

\subsection*{Baxter-King Filter (BK)}

The BK filter (Baxter and King 1999) extracts a \emph{frequency-band} component:
it retains periodic variations between $[p_L, p_H]$ periods and eliminates
trend and high frequencies. For business cycles:
$p_L = 6$, $p_H = 32$ quarters ($1{,}5$ to $8$ years).

\subsection*{Holt-Winters Exponential Smoothing (ETS)}

The ETS (\emph{Error-Trend-Seasonal}) model by Holt and Winters combines:
level, trend and seasonality equations updated recursively with
exponentially declining weights. Parameters $(\alpha, \beta, \gamma)$
control the speed of adaptation of each component.

% ─────────────────────────────────────────────────────────────────────────────
\section{DGP Verification}

DGP with linear trend, annual cycle and semi-annual seasonality, $T = 120$ months.

\begin{lstlisting}[caption={Series with trend, cycle and seasonality}]
seed(42)
generate df t       = _n
generate df trend   = 0.05 * t
generate df cycle   = 0.5 * sin(t * 3.14159 / 12)
generate df season  = 0.3 * sin(t * 3.14159 / 6)
generate df noise   = (rnormal() - 0.5) * 0.3
generate df y       = 10.0 + trend + cycle + season + noise

hprescott(df, y, lambda=1600)
display df

baxter_king(df, y, low=6, high=32, k=12)
display df

let m_hw = holtwinters(df, y, trend=add, seasonal=add, period=12)
display m_hw
\end{lstlisting}

\subsection*{Hodrick-Prescott}

\begin{lstlisting}[language={}, basicstyle=\ttfamily\small,
  frame=none, numbers=none, backgroundcolor=\color{codbg},
  xleftmargin=0pt, xrightmargin=0pt, breaklines=false]
hpfilter: λ=1600  →  y_trend and y_cycle added to df
  t=1:  y=10.34  y_trend=10.56  y_cycle=-0.22
  t=2:  y=10.62  y_trend=10.56  y_cycle= 0.06
  ...
\end{lstlisting}

The HP adds columns \texttt{y\_trend} (smoothed trend) and \texttt{y\_cycle}
(cycle = residual).

\subsection*{Baxter-King}

The BK adds \texttt{y\_cycle} with the frequency-band component.
The first and last $k = 12$ periods are set to \texttt{NaN} — the cost
of two-sided filters.

\subsection*{Holt-Winters ETS}

\begin{lstlisting}[language={}, basicstyle=\ttfamily\small,
  frame=none, numbers=none, backgroundcolor=\color{codbg},
  xleftmargin=0pt, xrightmargin=0pt, breaklines=false]
======================== ETS(A,A,A) ========================
Observations:    120
Alpha:        0.8000     (level)
Beta:         0.0500     (trend)
Gamma:        0.6000     (seasonality)
SSE:            5.3570
AIC:          -339.09
BIC:          -291.70
============================================================
\end{lstlisting}

The ETS(A,A,A) model — additive errors, additive trend, additive seasonality —
fits the smoothing parameters $(\alpha, \beta, \gamma)$
minimizing SSE. High $\alpha$ (0{,}80) indicates fast level response.

% ─────────────────────────────────────────────────────────────────────────────
\section{Syntax}

\begin{lstlisting}
hprescott(df, var, lambda=1600)              // adds var_trend and var_cycle
baxter_king(df, var, low=6, high=32, k=12)  // adds var_cycle
holtwinters(df, var, trend=add, seasonal=add, period=12)
ets(df, var, trend=add, seasonal=add, period=12)  // alias
\end{lstlisting}

\begin{center}
\begin{tabular}{lll}
\toprule
\textbf{Filter} & \textbf{Output} & \textbf{Typical use} \\
\midrule
HP              & trend + cycle         & quarterly macro \\
BK              & band-pass cycle       & business cycle analysis \\
Holt-Winters    & ETS model + forecast  & short-term forecasting \\
\bottomrule
\end{tabular}
\end{center}

\begin{nota}
  The HP filter on end-of-sample data suffers from the
  \emph{end-point problem}: trend and cycle are revised as new
  data arrives. The BK avoids this problem by using a symmetric window
  (cost: loses $k$ observations at each end).
\end{nota}
