\chapter{Advanced Panel II}
\label{cap:painel2}
% ── index ───────────────────────────────────────────────────────
\index{Mundlak, estimator of|textbf}
\index{mundlak@\texttt{mundlak()}|textbf}
\index{correlated random effects|see{Mundlak, estimator of}}
\index{CRE|see{Mundlak, estimator of}}
\index{xtgls@\texttt{xtgls()}|textbf}
\index{mixed model|textbf}
\index{mixedlm@\texttt{mixedlm()}|textbf}
\index{conditional logit|textbf}
\index{clogit@\texttt{clogit()}|textbf}
\index{REML}

% ─────────────────────────────────────────────────────────────────────────────
\section{Beyond basic FE and RE}

The estimators from ch.~\ref{cap:painel_avancado} (AB-GMM, PCSE, GEE) deal
with dynamics and heteroscedasticity. Here we cover four complementary approaches:

\begin{itemize}
  \item \textbf{FE} — reference; standard fixed effects (within estimator)
  \item \textbf{Mundlak (CRE)} — Correlated Random Effects; tests RE vs FE via $\gamma$
  \item \textbf{xtgls} — panel GLS; flexible regarding the structure of $\Sigma$
  \item \textbf{mixedlm} — mixed model with random intercepts
  \item \textbf{clogit} — conditional logit for binary panel variables
\end{itemize}

\subsection*{Mundlak Estimator (CRE)}

Mundlak (1978) shows that the RE estimator is equivalent to FE if:
\[
  \alpha_i = \gamma'\,\bar{x}_i + u_i \quad u_i \perp x_{it}
\]
Regresses $y_{it}$ on $x_{it}$ \emph{and} $\bar{x}_i$ (group mean).
If $\hat{\gamma} \neq 0$, the fixed effect is necessary.

\subsection*{Panel GLS (xtgls)}

Panel GLS specifies a structure for the error covariance across
entities. With \texttt{panels=hetero}, each panel has its own variance
$\sigma_i^2$, allowing heteroscedasticity across panels.

\subsection*{Mixed Model (mixedlm)}

The linear mixed model specifies:
\[
  y_{it} = \beta x_{it} + u_i + \varepsilon_{it} \quad u_i \sim N(0, \sigma_u^2)
\]
where $u_i$ is a random intercept effect per entity. It differs from standard RE
by using REML and integrating over $u_i$ via maximum likelihood.

% ─────────────────────────────────────────────────────────────────────────────
\section{DGP Verification}

DGP: endogenous static panel. $y = 1 + 2x + \alpha_i + \varepsilon$;
$\alpha_i = 0{,}05 \cdot i$ correlated with $x$ (via $z$).
$N = 50$ entities, $T = 20$ periods.

\begin{lstlisting}[caption={Panel II: FE, Mundlak, xtgls, mixedlm, clogit}]
seed(42)
generate df unit  = (_n - 1) % 50 + 1
generate df t     = (_n - (_n - 1) % 50 - 1) / 50 + 1
generate df alpha = unit * 0.05
generate df z     = (rnormal() - 0.5) * 1.7321
generate df x     = 0.5 * z + alpha + (rnormal() - 0.5)
generate df y     = 1.0 + 2.0 * x + alpha + e

let m_fe   = fe(y ~ x, df, id=unit)
let m_mund = mundlak(df, y ~ x, id="unit")
let m_gls  = xtgls(y ~ x, df, id=unit, time=t)
let m_mix  = mixedlm(y ~ x, df, id="unit")
generate df yb = (y > 2.5)
let m_cl   = clogit(yb ~ x, df, group="unit")
\end{lstlisting}

\begin{lstlisting}[language={}, basicstyle=\ttfamily\small, frame=none,
  numbers=none, backgroundcolor=\color{codbg}]
FE (within):        x = 2.041 ***    (true: 2.0) ✓
Mundlak:            x = 2.040,  γ(x̄) = 0.939 ***  → rejects RE
xtgls (hetero):     x = 2.768 ***    (biased without demeaning)
mixedlm:            x = 2.134 ***    (random intercept: σ²_u = 0.388)
clogit (yb ~ x):    x = 1.397 ***    (groups with no variation dropped)
\end{lstlisting}

\subsection*{Mundlak Test}

The test $H_0: \gamma = 0$ (means uncorrelated with $x$) rejects with
$F = 419{,}96$, $p < 0{,}001$. This confirms that $x$ is endogenous — the
RE estimator would be inconsistent and FE should be used.

The xtgls without FE removal overestimates $\beta$ because it captures part of the
individual effect. The mixedlm stays close to FE because it treats $\alpha_i$ as a
normal random effect.

% ─────────────────────────────────────────────────────────────────────────────
\section{Syntax}

\begin{lstlisting}
fe(y ~ x, df, id=unit)                    // Fixed Effects (within)
mundlak(df, y ~ x, id="unit")             // CRE: includes group means
xtgls(y ~ x, df, id=unit, time=t)         // panel GLS
xtgls(y ~ x, df, id=unit, time=t, panels="corr")  // correlated Σ
mixedlm(y ~ x, df, id="unit")             // mixed model (REML)
clogit(yb ~ x, df, group="unit")          // conditional logit
\end{lstlisting}

\begin{nota}
  In clogit, groups where $y$ does not vary (all 0 or all 1) are
  automatically dropped because they do not contribute to the conditional
  likelihood. This reduces the effective sample, but is theoretically correct.
\end{nota}
